---
title: Data Analysis - Exploratory Graphics
author: Bryan
date: '2018-07-22'
slug: data-analysis-exploratory-graphics
categories:
  - Data Analysis
  - Exploratory Data Analysis
  - R Programming
tags:
  - boxplots
  - Data Exploration
  - data preprocessing
  - data wrangling
---

### Exploratory Graphics in R

The focus of this analysis is on the ways to graph data set variables. The first section uses the EPA data set retrieved from: https://github.com/jtleek/modules/blob/master/04_ExploratoryAnalysis/exploratoryGraphs/data/avgpm25.csv the data was imported into a csv file and then used for this first section. Code notes for explaination. 

The first part uses the RStudio base functionality to plot graphics then we use maps, lattice, and ggplot2 to compare and contrast. ggplot2 and lattice produce nice graphics quickly, while maps has state maps with county delineations among other options. 

```{r loadLibs, message=FALSE, warning=FALSE}
library(tidyverse)
library(maps) #load maps package
library(lattice) # load lattice package
library(MASS) #load MASS package for cats data
```


```{r Loadata, message=FALSE, warning=FALSE}
avgpm25 <- read.csv("https://raw.githubusercontent.com/jtleek/modules/master/04_ExploratoryAnalysis/exploratoryGraphs/data/avgpm25.csv", row.names=NULL) #load data
str(avgpm25) #view structure of the data
avgpm25$fips <- as.character(avgpm25$fips) #change fips to a character variable. 
str(avgpm25) #check again
fivenum(avgpm25$pm25) #direct way to look at min, max, median, 1st & 3rd quartiles 25%/75%.
summary(avgpm25$pm25) #note same with addition of the mean
```

```{r}
boxplot(avgpm25$pm25, col="gold3") 
```

Dplyr to filter the data and then applied to the maps showing California and the counties where the readings are above 15. 

```{r}
#plot the five numbers from above and shows outlier dots
filter(avgpm25, pm25 > 15) #look at the points (dots) above 15
map("county", "California") #plot map of CA with county boundaries
with(filter(avgpm25, pm25 >15), points(longitude, latitude)) #location of the above 15 using lat/long
hist(avgpm25$pm25, col = "magenta")#look at distribution of the pm25 variable
hist(avgpm25$pm25, col="orange") #different color, see outliers
rug(avgpm25$pm25) #shows the data points at the bottom of the graph, darker is more clustered
hist(avgpm25$pm25, col="yellow", breaks=100)#add more detail to the bars to show data
rug(avgpm25$pm25) # add the data points to the bottom. 
boxplot(avgpm25$pm25, col="red") #back to the box plot
abline(h=12, lwd=4) #add a line at 12 were the 3rd quartile ends
hist(avgpm25$pm25, col="grey")# same kind of thing with the histogram
abline(v=12, lwd=3) #dark line added to show the end of the 75%.
abline(v=median(avgpm25$pm25), col= "cyan", lwd=4) #lets see where the median is in relation
table(avgpm25$region) %>% #dplyr table with a new color showing the distribution between regions
  barplot(col= "wheat")
boxplot(pm25 ~ region, data=avgpm25, col="green") #looking at the 5 numbers for region, outliers seen
par(mfrow = c(2,1), mar = c(4,4,2,1)) #set to plot 1 columns 2 row of graphs.
hist(subset(avgpm25, region== "east")$pm25, col="orange") #first histogram
hist(subset(avgpm25, region== "west")$pm25, col="purple") #second histogram together similar to boxplots comparision, skewing and outliers.
with(avgpm25, plot(latitude, pm25)) #basic scatterplot for latitude and pm25, shows group density and outliers
abline(h=12, lwd=2, lty=2) #adding the 12 for the end of 75% quartile
with(avgpm25, plot(latitude, pm25, col=region)) #color to contrast regions
abline(h=12, lwd=2,lty=2) #show where 12 is again.
levels(avgpm25$region) #checking the region levels same as before
table(avgpm25$region) #check the distribution in numeric form, maybe do this first before graphing.
par(mfrow=c(1,2), mar=c(5,4,2,1)) #setup plot format again differ margins
with(subset(avgpm25, region == "west"), plot(latitude, pm25, main="West"))
with(subset(avgpm25, region == "west"), plot(latitude, pm25, main="East")) #compare side by side scatter plots.

xyplot(pm25 ~ latitude | region, data=avgpm25) #very fast plot to compare Regions

qplot(latitude, pm25, data=avgpm25, facets= .~ region)#a fast comparision plot of regions
#dev.off() #reset par()
data("airquality")#load data
with(airquality, {plot(Temp, Ozone) #adds a trend line to the scatter plot for clarity 
        lines(loess.smooth(Temp, Ozone)) #temp to ozone correlation
})
data(cars) #load data
with(cars, {plot(speed, dist)
        title("Speed vs. Stopping distance") #describe what the graph is about
})
state <- data.frame(state.x77, region=state.region) #create a data frame life expec to income broken down by region
xyplot(Life.Exp ~ Income | region, data=state, layout=c(4,1)) #using lattice package shows 4 categorical regions
data(mpg) #load data
qplot(displ, hwy, data=mpg) #ggplot2, very fast plot of miles per gallon and engine displacement. 

data(cats) #load data
head(cats) #view head of data
with(cats, {plot(Hwt~Bwt)
        lines(lowess(Hwt ~ Bwt), col="blue", lwd=4)}) #plot data with the regression line
```


## Iris Data Set

This was done after the other work was finished and shows same analysis to the iris data set that comes with RStudio. 


```{r}
data("iris")
str(iris) #view structure of the data
fivenum(iris$Sepal.Length) #direct way to look at min, max, median, 1st & 3rd quartiles 25%/75%.
fivenum(iris$Sepal.Width) #direct way to look at min, max, median, 1st & 3rd quartiles 25%/75%.
fivenum(iris$Petal.Length) #direct way to look at min, max, median, 1st & 3rd quartiles 25%/75%.
fivenum(iris$Petal.Width) #direct way to look at min, max, median, 1st & 3rd quartiles 25%/75%.
summary(iris) #note same with addition of the mean
boxplot(iris$Sepal.Length, col="magenta") 
boxplot(iris$Sepal.Width, col="green") 
boxplot(iris$Petal.Length, col="orange") 
boxplot(iris$Petal.Width, col="blue") 

filter(iris, Sepal.Width > 4) #look at the points (dots) above 4
hist(iris$Sepal.Length, col = "magenta")#look at distribution of the variable
rug(iris$Sepal.Length) #shows the data points at the bottom of the graph, darker is more clustered
hist(iris$Sepal.Width, col = "green")#look at distribution of the variable
rug(iris$Sepal.Width) #shows the data points at the bottom of the graph, darker is more clustered
hist(iris$Petal.Length, col = "orange")#look at distribution of the variable
rug(iris$Petal.Length) #shows the data points at the bottom of the graph, darker is more clustered
hist(iris$Petal.Width, col = "blue")#look at distribution of the variable
rug(iris$Petal.Width) #shows the data points at the bottom of the graph, darker is more clustered
hist(iris$Sepal.Length, col="magenta", breaks=100)#add more detail to the bars to show data
hist(iris$Sepal.Width, col="green", breaks=100)#add more detail to the bars to show data
hist(iris$Petal.Length, col="orange", breaks=100)#add more detail to the bars to show data
hist(iris$Petal.Width, col="blue", breaks=100)#add more detail to the bars to show data
boxplot(iris$Sepal.Length, col="magenta") #back to the box plot
abline(h=6.5, lwd=4) #add a line at 12 were the 3rd quartile ends
boxplot(iris$Sepal.Width, col="green") #back to the box plot
abline(h=3.5, lwd=4) #add a line at 12 were the 3rd quartile ends
boxplot(iris$Petal.Length, col="orange") #back to the box plot
abline(h=5.1, lwd=4) #add a line at 12 were the 3rd quartile ends
boxplot(iris$Petal.Width, col="blue") #back to the box plot
abline(h=1.8, lwd=4) #add a line at 12 were the 3rd quartile ends
hist(iris$Sepal.Length, col="grey")# same kind of thing with the histogram
abline(v=6.5, lwd=3) #dark line added to show the end of the 75%.
abline(v=median(iris$Sepal.Length), col= "cyan", lwd=4) #lets see where the median is in relation
table(iris$Species) %>% #dplyr table with a new color showing the distribution between Species
  barplot(col= "wheat")
boxplot(Sepal.Length ~ Species, data=iris, col="magenta") #looking at the 5 numbers 
boxplot(Sepal.Width ~ Species, data=iris, col="green") #looking at the 5 numbers 
boxplot(Petal.Length ~ Species, data=iris, col="orange") #looking at the 5 numbers 
boxplot(Petal.Width ~ Species, data=iris, col="blue") #looking at the 5 numbers
par(mfrow = c(2,2), mar = c(4,4,2,2)) #set to plot 1 columns 2 row of graphs.
hist(subset(iris, Species == "setosa")$Sepal.Length, col="magenta") #first histogram
hist(subset(iris, Species == "setosa")$Sepal.Width, col="green")
hist(subset(iris, Species == "setosa")$Petal.Length, col="orange")
hist(subset(iris, Species == "setosa")$Petal.Width, col="blue")
with(iris, plot(Sepal.Length, Sepal.Width, col = Sepal.Width)) #basic scatterplot
abline(h=3.5, lwd=2, lty=2) #adding the 3.5 for the end of 75% quartile
iris$Species <- as.factor(iris$Species)
levels(iris$Species) #checking the region levels same as before
table(iris$Species) #check the distribution in numeric form, maybe do this first before graphing.
par(mfrow=c(2,2), mar=c(5,4,2,1)) #setup plot format again differ margins
with(subset(iris, Species == "setosa"), plot(Sepal.Length, Sepal.Width, main="Setosa"))
with(subset(iris, Species == "versicolor"), plot(Sepal.Length, Sepal.Width, main="Versicolor")) #compare side by side scatter plots.
with(subset(iris, Species == "virginica"), plot(Sepal.Length, Sepal.Width, main="Virginica"))
with(subset(iris, Species == "setosa"), plot(Petal.Length, Petal.Width, main="Setosa"))
with(subset(iris, Species == "versicolor"), plot(Petal.Length, Petal.Width, main="Versicolor")) #compare side by side scatter plots.
with(subset(iris, Species == "virginica"), plot(Petal.Length, Petal.Width, main="Virginica"))
with(subset(iris, Species == "setosa"), plot(Sepal.Length, Petal.Width, main="Setosa"))
with(subset(iris, Species == "versicolor"), plot(Sepal.Length, Petal.Length, main="Versicolor")) #compare side by side scatter plots.

xyplot(Sepal.Width ~ Sepal.Length | Species, data=iris) #very fast plot to compare 
xyplot(Petal.Width ~ Petal.Length | Species, data=iris) #very fast plot to compare 
xyplot(Sepal.Width ~ Petal.Width | Species, data=iris)
xyplot(Sepal.Length ~ Petal.Length | Species, data=iris)

qplot(Sepal.Length, Sepal.Width, data=iris, facets= .~ Species)#a fast comparision plot of
qplot(Petal.Length, Petal.Width, data=iris, facets= .~ Species)
qplot(Sepal.Length, Petal.Width, data=iris, facets= .~ Species)
qplot(Sepal.Length, Petal.Length, data=iris, facets= .~ Species)
qplot(Sepal.Width, Petal.Width, data=iris, facets= .~ Species)
qplot(Sepal.Width, Petal.Length, data=iris, facets= .~ Species)
```
