---
title: Simple data exploration with base R functions
author: 'Bryan '
date: '2018-07-02'
slug: simple-data-exploration-with-base-r-functions
categories:
  - Exploratory Data Analysis
  - Data Analysis
  - R Programming
tags:
  - data collection
  - data manipulation
---

#### Exploring the data with base R functions

The purpose of this post is to show the basic fucntions in R to start a data analysis project. I look at the iris dataset that comes with Rstudio shown below.
This briefly covers `head()`, `str()`, `attributes()`, `summary()`, `dim()`, `names()`, indexing [], `table()`, and `plot()` functions that come with base R. Simple to use. 

#### Load the iris dataset

```{r}
data("iris") 
```

#### head()

Look at the iris dataset with the `head()` function and lists the first five rows. 

```{r}
head(iris)  
```

#### str()

View the structure of the dataset. 150 objects and 5 variables.

```{r}
str(iris) 
```

#### attributes()

View the object attributes.

```{r}
attributes(iris)
```

#### summary()

Here is the summary statistics. Min/Max, Mean/Median, Quantiles, also a place to see if there is missing data (NA) which there is none in this iris data set. 

```{r}
summary(iris) 
```

#### dim()

#### names()

We can use the `dim()` function, `names()` function to view this information. Which `str()` did for use above more completely.  

```{r}
dim(iris) 
names(iris)
```

#### index [a,b
]
Below is the view of rows and columns and shows the proper syntax for doing this. [12,2] is the index for row 12 column 2 of the data frame. 

```{r}
iris[12, 2]
```

Or you can view row 12 in total by indexing [12, ]

```{r}
iris[12, ] 
```

You can index into the middle and view rows 10 through 15. 

```{r}
iris[10:15, ] 
```

Here is the way to view column 5. 

```{r}
iris[ ,5]
```

View rows 1, 5, 19, 15 and the respective column variables.

```{r}
iris[c(1,5,10,15), ] 
```

Or view rows 1, 5, 10, 15 in column 1 only. Note by default this converts the column to a row in R.

```{r}
iris[c(1,5,10,15), 1]
```

#### table()

Here we can view the data in a table and basic plot. The tables show the three categorical variables (factors in R) and Petal.Width and Sepal.Width variables (integers). We can see a pattern develop with each Species/Petal.Width and Species/Sepal.Width when plotted.   

```{r}
table(iris$Species, iris$Petal.Width)
petwidth <- table(iris$Species, iris$Petal.Width)
```

#### plot()

The plot is multi-dimensional.
```{r}
barplot(petwidth, main = "Petal Width")
plot(petwidth, main = "Petal Width", col = "green")
```

#### table()

```{r}
table(iris$Species, iris$Sepal.Width)
sepwidth <- table(iris$Species, iris$Sepal.Width)

```

#### plot()

```{r}
barplot(sepwidth, main = "Sepal Width")
plot(sepwidth, main = "Sepal Width", col = "orange")
```
